Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2015-04-12 slides/esop-termination.tex
More esop slides

@@

@@

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}

@@

In our logical encoding, we model \textcolor{green}{G} as a predicate $\textcolor{green}{G}(x)$. \textcolor{red}{P} and \textcolor{blue}{B} are modelled as transition
relations $\textcolor{red}{P}(x, x')$ and $\textcolor{blue}{B}(x, x')$.

\begin{frame}[fragile]{Loop Encoding Example}

\begin{columns}[c]
 \column{.5\textwidth}

 \begin{center}
 \begin{tabular}{c}

 \begin{lstlisting}[language=C,basicstyle=\normalsize,mathescape]
<@\textcolor{red}{x = 0}@>;

while (<@\textcolor{green}{x $<$ 100}@>) {
  <@\textcolor{blue}{x++}@>;
}
 \end{lstlisting}
 \end{tabular}
 \end{center}

 \column{.5\textwidth}
\begin{eqnarray*}
 \textcolor{red}{P} = & \{ (x, x') \mid x' = 0 \} \\
 \textcolor{green}{G} = & \{ x \mid x < 100 \} \\
 \textcolor{blue}{B} = & \{ (x, x') \mid x' = x + 1 \}
\end{eqnarray*}
\end{columns}

@@

\begin{frame}[fragile]{Ranking Functions}
 To prove that a loop $(P, G, B)$ terminates, we find a \emph{ranking function} $R$ which satisfies the following criteria:

 \begin{eqnarray*}
  \forall x . G(x) \Rightarrow & R(x) > 0 \\
  \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
 \end{eqnarray*}

\begin{exampleblock}{Example}
\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x < 100) {
  x++;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 $R(x) = 100 - x$
\end{columns}
\end{center}
\end{exampleblock}

\begin{frame}[fragile]{Conditional Ranking Functions}
 A loop may not terminate starting in some initial state.  However, the prefix may guarantee
 that the loop does terminate.

 To prove that such a loop terminates, we must find a \emph{supporting invariant} $I$ along with the
 ranking function $R$:

 \begin{eqnarray*}
  \forall x, x' . P(x, x') \Rightarrow & I(x') \\
  \forall x, x' . I(x) \wedge B(x, x)' \Rightarrow & I(x') \\
  \forall x . I(x) \wedge G(x) \Rightarrow & R(x) > 0 \\
  \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
 \end{eqnarray*}
\end{frame}

\begin{frame}[fragile]{Conditional Ranking Function Example}
\begin{exampleblock}{Example}
\begin{center}
\begin{columns}[T]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
y = 1;
while (x < 100) {
  x += y;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
\begin{gather*}
  I =  \{ (x, y) \mid y = 1 \} \\
  R(x, y) =  100 - x
\end{gather*}
\end{columns}
\end{center}
\end{exampleblock}
\end{frame}

\begin{frame}{Second-Order Termination}

\begin{block}{Conditional Termination Formula [CT]}
 \[
 \exists \red{I}, \blue{R} .  \forall x, x' . \left ( \begin{array}{rl}
   P(x, x') \Rightarrow & \red{I}(x') \, \wedge \\
   \red{I}(x) \wedge G(x) \Rightarrow & \blue{R}(x) > 0 \, \wedge \\
   \red{I}(x) \wedge G(x) \wedge B(x, x') \Rightarrow & \blue{R}(x) > \blue{R}(x') \wedge \red{I}(x')
   \end{array} \right )
 \]
 \end{block}

This formula is satisfiable iff there exists a ranking function $\blue{R}$ and supporting invariant $\red{I}$,
i.e. iff the loop $(P, G, B)$ terminates.
